### 2025-11-08 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
2025NSF: integrated most comments from Alejandro

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal,HayashiQuantumInfo,majtey2005jensen}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropic structure (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and the convex structure are readily available, to our knowledge, the correct axioms for the entropic structure still need to be identified. Moreover, the interplay between the topology, the convex structure and the entropic structure opens additional mathematical questions which have not been investigated.

The goal of this proposed work is to develop a common axiomatic basis for the information-theoretic structure underlying all physical theories, classical and quantum. The immediate aim is to generalize results of both classical and quantum information theory to a common, more abstract theory that can also handle the infinite-dimensional case. Longer term, we aim to use the axioms developed here as part of a larger effort to develop a unified base theory for physical theories with a clean mathematical foundation.

Note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, the biggest problem is understanding how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we already have found the solution. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given precisely when the two ensembles are orthogonal with respect to the inner product of the respective space. In the classical case, this means that the measures have disjoint support; in the quantum case, it means that the eigenspaces of non-zero eigenvalues are orthogonal. In other words, characterizing the interaction of the entropy function with the topology and the convex structure is enough to characterize the entropic structure.  That is, the form of $I(p)$---the Shannon entropy formula~\cite{Shannon}---can be characterized, and since in both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case.

Another important insight from our preliminary work is that the entropy is not only responsible for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. That is, the entropic structure of the ensemble space is equivalent to the geometric structure of the state space.

As we saw in classical mechanics, the symplectic form gives us the Liouville measure which in turn is used to define the entropy. Therefore, the entropy of an ensemble is defined from the geometry of the state space. However, note that the entropy for a uniform distribution $\rho_U$ over a set $U$ is given by $S(\rho_U) = \log \mu(U)$. Therefore, if we are given the entropy for all probability measures, we can recover the Liouville measure by setting $\mu(U) = 2^{S(\rho_U)}$. While this does not recover the symplectic form directly, note that the symplectic form can also be used to define the entropy of the marginal distributions along each degree of freedom. Around a point, in fact, we can define Darboux coordinates and define joint probability distributions that are the product of independent marginals along each pair of position and momentum. One finds a similar relationship between the entropy of uniform marginal distributions and the integral of the symplectic form over the support of the marginal. From these, we believe we can recover the symplectic form.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has von Neumann entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi|\psi)}}{2} \right)$ where $I$, as before, is the Shannon entropy for two elements. As before, the entropy is defined in terms of the inner product. However, the relationship between $S(\rho_{\psi\phi})$ and $p(\phi|\psi)$ is invertible. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the geometry of the state space.

@@

The central focus is the intuitive notion of statistical ensemble (i.e.~the collection of all outputs of a preparation procedure), as this seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the mathematics, as we have seen, the spaces of statistical ensembles of classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that any physical theory must \emph{at least} posit a space of ensembles. This provides us with a strong conceptual starting point, and properties of ensembles are what need to be specified through axioms.

@@

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space.

Surprisingly, topological convex spaces do not seem to have been studied, and this structure alone presents interesting sets of questions.  For example, it is established that a convex space embeds into a vector space if and only if it is cancellative, that is, if $p \ens[a] + \bar{p} \ens = p \ens[b] + \bar{p} \ens$ for some $p \in (0,1)$ implies $\ens[a] = \ens[b]$. The question now is what are necessary and sufficient conditions for a topological convex space to embed continuously into a topological vector space? Can we have, for example, a Hausdorff second-countable topological convex space that is not a continuous embedding of a topological vector space? Topological vector spaces present a number of nice results which may or may not hold in topological convex spaces. For example, any $\mathsf{T}_1$ second-countable topological vector space is metrizable. Does that same result hold for topological convex spaces? Another interesting aspect is the notion of closure. In a topological vector space, the topological closure of a convex set is convex. We found that this is true in topological convex spaces as well. Additionally, there is another type of closure that may be more interesting physically. We can define the $\sigma$-hull of a set $A$ as the set of possible infinite convex combinations, which can also be shown to be a convex set. This will be a superset of the hull of $A$, because it will include all finite convex combinations, and it will be a subset of the closed hull of $A$, as all infinite convex combinations are topological limits. Moreover, it can be shown that, in some cases, the hull, the $\sigma$-hull and the closed hull are all different. From the physics side, the $\sigma$-hull may be a more physically meaningful closure as it represents the true limit process of statistical mixing, the operation the axiom is characterizing. The subtle issue is that convex combinations always return elements ``inside'' the set, while linear combinations can also return the ``edges.'' Therefore, there are already interesting topics to study even without imposing the entropic structure.

@@

In the finite-dimensional case, and if the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. Let $V_{\ens[a]}$ be the embedding vector space with $\ens[a] \in \Ens$ as origin. This will be a differentiable manifold. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold~\cite{hessianmanifoldsShima1997,hessianmanifoldsliu2025}. In classical and quantum mechanics it recovers the Fisher information metric (see e.g.~Refs.~\cite{facchi2010,AmariInfoGeom}). The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

Another goal is to recover the geometric structure of the state spaces of both classical and quantum mechanics from the entropic structure, to formally show the equivalence. For quantum mechanics, one can simply define the Born rule from the entropy as we have seen before. For classical mechanics, the situation is a bit more complex as we need to recover first the Liouville measure. This can be done by noting two things: for a uniform distribution with support $U \subseteq X$, $\mu(U) = 2^{S(\rho_U)}$, and the uniform distribution over $U$ has the highest entropy of all distributions with support $U$. Given a set $A \subseteq \Ens$ of ensembles, we can then define the state capacity as the supremum of the exponential of the hull of $A$, which physically means the highest entropy achievable with a statistical mixture of elements of $A$. This set function has nice properties: it is non-negative, monotonic, subadditive in general and additive on orthogonal sets of ensembles. Note that if $A_U \subseteq \Ens$ is the subspace of all probability measures with support $U \subseteq X$, the state capacity of $A_U$ recovers the Liouville measure of $U$. Moreover, $A_U$ is orthogonal to $A_V$ if and only if $U$ and $V$ are disjoint. This means that the state capacity restricted to the lattice of those subspaces is an additive measure. Interestingly enough, one can follow the same construction in the quantum case, and the measure will return the dimensionality of the subspace. Concerning non-additive measures, there are additional interesting questions as one can create a non-additive version of probability measures, which may point to a non-additive version of probability theory. We have already looked at this problem, and, unfortunately, the standard techniques that have been developed in that area are not sufficient to handle the cases of interest to us within physical theories.

>>>>

While the structures of the state spaces of classical and quantum mechanics are different, their spaces of ensembles have a rather similar structure: they both consist of a bounded convex subset of a metrizable and second-countable locally convex topological vector space, where the metric can be derived from the entropy through the square root of the Jensen-Shannon divergence $JSD = S\left(\frac{1}{2} \ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2}S(\ens[a]) + \frac{1}{2}S(\ens[b])\right)$~\cite{Jensen-ShannonOriginal,HayashiQuantumInfo,majtey2005jensen}. The JSD is, in both cases, the square of a distance function and can be interpreted as how much the average entropy increases under an equal mixture of two ensembles. There are, therefore, three common structures at play: a topology (which physically represents experimental verifiability), a convex structure (which physically represents the ability to make statistical mixtures) and an entropy function (which physically represents the variability of the elements within each ensemble). While the axioms for the topology and for convex spaces(CITE) are readily available, to our knowledge, the correct axioms for an entropy function and its required compatibility condition with the other two structure still need to be identified.

The goal of this proposed work is to develop a common axiomatic basis for the information-theoretic structure underlying all physical theories, classical and quantum, which we will call \textbf{entropic structure}.  The immediate aim is to generalize results of both classical and quantum information theory to a common, more abstract theory that can also handle the infinite-dimensional case. Longer term, we aim to use the axioms developed here as part of a larger effort to develop a unified base theory for physical theories with a clean mathematical foundation.

Before looking at our preliminary proposal for the axiomatic, let us develop some intuition about the structure we need to generalize. Note that in all above three cases the definition of the entropy is tightly bound to the definition of the space. Therefore, the biggest problem is understanding how one can give a single characterization that is general enough to encompass the three cases, while constrained enough to recover the correct expression in each case. Fortunately, in our preliminary work we have found the right approach. In all cases, the entropy possesses an upper bound to its increase under convex combination: $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p) + p S(\ens[a]) + \bar{p} S(\ens[b])$ where $\bar{p} = 1-p$ and $I(p) = -p \log p - \bar{p}\log \bar{p}$. This maximum increase is given, in the classical case, when the two measures have disjoint support; in the quantum case, it is given when the eigenspaces of non-zero eigenvalues are orthogonal. Our approach, then, is to proceed in the opposite way: impose a generic entropy function to have an upper bound and enough properties to show that said upper bound must correspond to the Shannon entropy formula~\cite{Shannon}. Since both classical and quantum mechanics ensembles can always be decomposed into orthogonal components, the different entropy formulas can be recovered in each case as a recursive expansion of the upper bound.

Another important insight from our preliminary work is that the entropy is not only responsible for the geometry of the ensemble space, but it is also intimately related to the geometry of the state space of both classical and quantum mechanics. The idea, then, is to recover to the geometric structure of the state space from it entropic structure.

For example, as we saw in classical mechanics, the symplectic form gives us the Liouville measure which in turn is used to define the entropy. Therefore, the entropy of an ensemble is defined from the geometry of the state space. However, note that the entropy for a uniform distribution $\rho_U$ over a set $U$ is given by $S(\rho_U) = \log \mu(U)$. Therefore, if we are given the entropy for all probability measures, we can recover the Liouville measure by setting $\mu(U) = 2^{S(\rho_U)}$. While this does not recover the symplectic form directly, note that the symplectic form can also be used to define the entropy of the marginal distributions along each degree of freedom. Around a point, in fact, we can define Darboux coordinates and define joint probability distributions that are the product of independent marginals along each pair of position and momentum. One finds a similar relationship between the entropy of uniform marginal distributions and the integral of the symplectic form over the support of the marginal. From these, we believe we can recover the symplectic form.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
as this is the cosine square of the distance function of the Fubini-Study metric. Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has von Neumann entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi|\psi)}}{2} \right)$ where $I$, as before, is the Shannon entropy for two elements. As before, the entropy is defined in terms of the inner product. However, the relationship between $S(\rho_{\psi\phi})$ and $p(\phi|\psi)$ is invertible. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the geometry.

@@

The central focus is the intuitive notion of statistical ensemble (i.e.~the collection of all outputs of a preparation procedure), as this seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the mathematics, as we have seen, the spaces of statistical ensembles of classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that any physical theory must \emph{at least} posit a space of ensembles. This provides us with a strong conceptual starting point, and properties of ensembles are what need to be specified through axioms. We will therefore call \textbf{ensemble space} a set endowed with the suitable topology, convex structure and entropic structure that generalizes the spaces of classical and quantum ensembles.

@@

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space. Given the analogy, we call \textbf{topological convex space} a convex space that is also a topological space for which the mixing operation is continuous.

Topological convex spaces have not been developed in the literature, and this structure alone presents interesting sets of questions.  For example, it is established that a convex space embeds into a vector space if and only if it is cancellative, that is, if $p \ens[a] + \bar{p} \ens = p \ens[b] + \bar{p} \ens$ for some $p \in (0,1)$ implies $\ens[a] = \ens[b]$. The question now is what are necessary and sufficient conditions for a topological convex space to embed continuously into a topological vector space? Can we have, for example, a Hausdorff second-countable topological convex space that is not a continuous embedding of a topological vector space? Topological vector spaces present a number of nice results which may or may not hold in topological convex spaces. For example, any $\mathsf{T}_1$ second-countable topological vector space is metrizable. Does that same result hold for topological convex spaces? Another interesting aspect is the notion of closure. In a topological vector space, the topological closure of a convex set is convex. We found that this is true in topological convex spaces as well. Additionally, there is another type of closure that may be more interesting physically. We can define the $\sigma$-hull of a set $A$ as the set of possible infinite convex combinations, which can also be shown to be a convex set. This will be a superset of the hull of $A$, because it will include all finite convex combinations, and it will be a subset of the closed hull of $A$, as all infinite convex combinations are topological limits. Moreover, it can be shown that, in some cases, the hull, the $\sigma$-hull and the closed hull are all different. From the physics side, the $\sigma$-hull may be a more physically meaningful closure as it represents the true limit process of statistical mixing, the operation the axiom is characterizing. The subtle issue is that convex combinations always return elements ``inside'' the set, while linear combinations can also return the ``edges.'' Therefore, there are already interesting topics to study even without imposing the entropic structure.

@@

Given an internal point $\ens[a] \in \Ens$ (i.e. for every $\ens[b] \in \Ens$ we can find $\ens[c] \in \Ens$ such that $\ens[a] = p \ens[b] + \bar{p} \ens[c]$ from some $p \in (0,1)$) we can construct an embedding vector space $V_{\ens[a]}$. If $V_{\ens[a]}$ is finite dimensional, we can show that $V_{\ens[a]}$ is a neighborhood of $\ens[a]$, which means that the interior of $\Ens$ is a manifold. If the entropy is twice differentiable with respect to the convex coefficients, the strict concavity of the entropy means that the Hessian is negative definite, which allows us to define a metric tensor. On a linear coordinate chart we can define $g_{\ens}( v, w ) = - \frac{\partial^2 S}{\partial \ens ^2}( v, w )$ where $v$ and $w$ are elements of the tangent space at $\ens \in \Ens \subset V_{\ens[a]}$. Since all Hessians are symmetric, $g_{\ens}$ will be bilinear, symmetric and positive definite: it is a metric tensor and $V_{\ens[a]}$ is a Hessian manifold~\cite{hessianmanifoldsShima1997,hessianmanifoldsliu2025}. In classical and quantum mechanics it recovers the Fisher information metric (see e.g.~Refs.~\cite{facchi2010,AmariInfoGeom}). The immediate goal, then, is to see how much of information geometry can be generalized to this case. If a locally convex topology can be established, the further goal is to extend this to the infinite-dimensional case.

Another construction allows us to generalize the notion of count of states given, in classical mechanics, by the Liouville measure. This can be done by noting two things: for a uniform distribution with support $U \subseteq X$, $\mu(U) = 2^{S(\rho_U)}$, and the uniform distribution over $U$ has the highest entropy of all distributions with support $U$. Given a set $A \subseteq \Ens$ of ensembles, we can then define the state capacity as the supremum of the exponential of the entropy within the hull of $A$, which physically means the highest entropy achievable with a statistical mixture of elements of $A$. This set function has nice properties: it is non-negative, monotonic, subadditive in general and additive on orthogonal sets of ensembles. Note that if $A_U \subseteq \Ens$ is the subspace of all probability measures with support $U \subseteq X$, the state capacity of $A_U$ recovers the Liouville measure of $U$. Moreover, $A_U$ is orthogonal to $A_V$ if and only if $U$ and $V$ are disjoint. This means that the state capacity restricted to the lattice of those subspaces is an additive measure. Interestingly enough, one can follow the same construction in the quantum case, and the measure will return the dimensionality of the subspace. Concerning non-additive measures, there are additional interesting questions as one can create a non-additive version of probability measures, which may point to a non-additive version of probability theory. We have already looked at this problem, and, unfortunately, the standard techniques that have been developed in that area are not sufficient to handle the cases of interest to us within physical theories.
